### 2021-05-31 docs/manuscript/manuscript.tex
Avoid line breaking of citations

Cells interact continuously with their environment and they must detect signals
from it. In many cases Gene Regulatory Networks (GRNs) interpret these signals
through gene interaction and produce a particular response. This information
processing function is of particular importance in developmental processes such
as those in which a group of cells differentiate in response to a signalling
molecule. Such molecules were referred to as morphogens by Turing in 1952
\cite{Turing1952} and posterior theoretical studies on patterning led to the
conceptualization of the French Flag Problem by Wolpert, who also stated
in this respect that a gradient of concentration of a morphogen could
trigger cell differentiation in a one-dimensional field of cells
\cite{Wolpert1969, sharpe_2019}.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80's in the developing embryos of
\textit{Drosophila melanogaster}
\cite{driever_gradient_1988, driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs \cite{affolter_decapentaplegic_2007},
as well as Sonic Hedgehog \cite{dessaud_dynamic_2010,cohen_theoretical_2014}
and Wnt \cite{raspopovic_digit_2014}. However, information processing by GRNs
has proven to be a complex process, and a major goal of developmental biology
is to understand mechanistically how positional information given by morphogens
is translated in spatial differentiation.\\

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, which are defined as patterns of
interconnections occurring in networks at numbers that are significantly higher
than those in randomized networks~\cite{Milo2002}. This novel approach, along
with mathematical modeling and computational systems biology methods have been
shown to be a powerful tool to study GRNs~\cite{kitano_computational_2002}, and
the predictions of studies using these methods have been validated by
experimental studies~\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}, and
some of them have even used synthetic GRNs to produce artificial cell
differentiation~\cite{Basu2005}, circadian gene
expression~\cite{atkinson_development_2003}, counting
devices~\cite{friedland_synthetic_2009} and systems that respond to light.
Moreover, the design and implementation of GRNs through synthetic biology is
emerging as a promising tool to study biological phenomena as pattern formation
\cite{santos-moreno_using_2019} and as a novel therapeutic tool with amazing
biomedical applications~\cite{karlsson_therapeutic_2012,
higashikuni_advancing_2017,abil_synthetic_2015,healy_genetic_2019,
kitada_programming_2018}.

Cotterell and Sharpe studied what kind of 3-node network topologies could
interact with a morphogen gradient and produce a stripe of gene expression
in a one-dimensional field of cells and found a variety of networks that
implemented at least 7 different mechanisms and had different levels of
complexity \cite{Cotterell2010}.

However, although several studies have analyzed the properties of GRNs that
process positional information, they have focused on particular natural networks
\cite{Schaerli2014,Jaeger2004}, in synthetic networks
\cite{Basu2005,Reinitz1995,Schaerli2018}, or have constrained the design of the
GRNs \cite{Elowitz2000}.

@@

Our model is a modification of the model used by Cotterell \& Sharpe
\cite{Cotterell2010} and proposed by Reinitz et al. \cite{Reinitz1995}. The
model is a dynamical system that describes the change in concentration of gene
product \emph{i} in time, as shown in Eq~\ref{ec:modelo-gral}.

@@

We adopted the same definition of neighborhood as Cotterell \& Sharpe
\cite{Cotterell2010}, in which two topologies were neighbors if the gain or
removal of any one interaction can transform one of the topologies into the
other.

@@

As a measure of network complexity, we used the complexity index based on
Shannon entropy proposed by Bonchev \& Rovray \cite{D.2005}. We calculated this
index using the following equation:
\begin{equation}
 I_{vd} = \sum_{i=1}^V a_i \cdot \log_{2} a_i,
\end{equation}
\noindent
being $a_i$ the degree of \emph{i}-th node and $V$ the number of nodes in the
network. The network descriptors were calculated using built-in functions from
Wolfram Mathematica software.\\

@@

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network \cite{Erdos1959}. This kind of network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}A). Nodes corresponding to topologies number one and
three showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:deg-dist}B). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
($r_S = 0.353752$, $p = 6.2933\times10^{-23} $).

@@

Besides topological robustness, we wanted to know if network topologies were
robust to perturbations in the values of the parameters. After performing the
perturbations to a GRN from each topology, we plotted the abundance of the
topology vs. its robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$)
(Fig~\ref{fig:ab-rob}A). However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif reported by Cotterell \&
Sharpe \cite{Cotterell2010}, the I3-FFL and the ``Overlapping domains'' motif
(which is a variation of the I3-FFL) also reported by Cotterell \& Sharpe (see
Supplementary Material).

@@

\begin{table}[!ht]
% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
 \caption{{\bf Network descriptors of neutral network of topologies.}}
 \begin{tabular}{|l|l|}
 \hline
 {\bf Network descriptor} & {\bf Value}\\ \thickhline
 Average vertex degree  & 3.85714 \\ \hline
 Clustering coefficient \cite{Watts1998} & 0.117097 \\ \hline
 Average path length    & 7.25261       \\ \hline
 Graph connectedness    & 0.00540974    \\ \hline
 Graph diameter*        & 25            \\ \hline
 Average path length*   & 9.0499        \\ \hline
 \end{tabular}
 \begin{flushleft} *These descriptors were calculated for the main connected
 subgraph of the network.
 \end{flushleft}
 \label{table1}
% \end{adjustwidth}
 \end{table}

@@

Robustness is an important feature of patterning mediated by morphogens
\cite{lo_robust_2015}. In our study, the most abundant topologies were found to
be part of highly connected clusters in the neutral network, and this fact
indicates that these topologies are both robust to quantitative changes in the
strength of interactions and diffusion and degradation parameters as well as to
qualitative changes in the topology itself. As robustness and evolvability are
not mutually exclusive properties \cite{wagner_robustness_2008}, it is likely
that these topologies are also evolvable.

Other studies have shown that Incoherent Feed Forward Loops are able to form
spatial stripes of gene expression \cite{ishihara_cross_2005}. In particular,
we found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \& Sharpe
\cite{Cotterell2010}, and has been found in transcriptional networks involved
in cell fate definition \cite{Li2019}. Moreover, our results indicate that this
is a robust motif that can be useful for construction of synthetic gene
regulatory networks that are exposed to stochastic fluctuations in the
environment and mutations.

@@

Although it has been experimentally demonstrated that the Incoherent Feed
Forward Loop type 2 (I2-FFL) is able to generate a striped gene expression
pattern, \cite{Schaerli2014, Basu2005}, in our study this motif was present in
only 10.92\% of the topologies. This fact may be because genetic algorithms
can not always find global minima but local minima or its surroundings, however,
it is also possible that the I2-FFL motif is not as robust as other motifs and
therefore be less represented.

We also found topologies with a similar topology to that of the “opposing
gradients” reported by Schaerli et al. \cite{Schaerli2018,Schaerli2014}.
Although the “opposing gradients” mechanism requires the constitutive expression
of two of the genes in the GRN, and our model does not take constitutive
expression into account, we realized that the GRNs that presented the “opposing
gradients” topology, had also commonly positive auto-regulation interactions on
these two nodes, supplying in this way the lack of constitutive expression.

@@

Although GRNs that produce a band of gene expression with only two nodes have
been described (e.g. the I-zero motif, \cite{Schaerli2014}) this is to our
knowledge the first time a GRN with a topology such as that of topology nine
(Fig \ref{fig:topol9})
and its expression mechanism are reported, increasing our knowledge of the
genotype-phenotype map in this kind of system, and it would be very interesting
to confirm experimentally the ability of this topology to generate the described
pattern and it would be interesting to study its feasibility and possible
implementation in synthetic gene regulatory networks.

@@

\paragraph*{S4 Fig.}
\label{S4_Fig}
{\bf Relevant subgraphs present in the set of GRNs.}
These are subgraphs that have been reported in previous studies as networks
involved in morphogenesis and development \cite{Cotterell2010, Schaerli2014,
Schaerli2018}. These were used to calculate the subgraph profile and the results
reported in \nameref{S1_Table}.

>>>>

Cells interact continuously with their environment and they must detect signals
from it. In many cases Gene Regulatory Networks (GRNs) interpret these signals
through gene interaction and produce a particular response. This information
processing function is of particular importance in developmental processes such
as those in which a group of cells differentiate in response to a signalling
molecule. Such molecules were referred to as morphogens by Turing in
1952~\cite{Turing1952} and posterior theoretical studies on patterning led to the
conceptualization of the French Flag Problem by Wolpert, who also stated
in this respect that a gradient of concentration of a morphogen could
trigger cell differentiation in a one-dimensional field of cells~\cite{Wolpert1969,
sharpe_2019}.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80's in the developing embryos of
\textit{Drosophila melanogaster}~\cite{driever_gradient_1988,
driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs~\cite{affolter_decapentaplegic_2007},
as well as Sonic Hedgehog~\cite{dessaud_dynamic_2010,cohen_theoretical_2014}
and Wnt~\cite{raspopovic_digit_2014}. However, information processing by GRNs
has proven to be a complex process, and a major goal of developmental biology
is to understand mechanistically how positional information given by morphogens
is translated in spatial differentiation.\\

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, which are defined as patterns of
interconnections occurring in networks at numbers that are significantly higher
than those in randomized networks~\cite{Milo2002}. This novel approach, along
with mathematical modeling and computational systems biology methods have been
shown to be a powerful tool to study GRNs~\cite{kitano_computational_2002}, and
the predictions of studies using these methods have been validated by
experimental studies~\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}, and
some of them have even used synthetic GRNs to produce artificial cell
differentiation~\cite{Basu2005}, circadian gene
expression~\cite{atkinson_development_2003}, counting
devices~\cite{friedland_synthetic_2009} and systems that respond to light.
Moreover, the design and implementation of GRNs through synthetic biology is
emerging as a promising tool to study biological phenomena as pattern
formation~\cite{santos-moreno_using_2019} and as a novel therapeutic tool with
interesting biomedical applications~\cite{karlsson_therapeutic_2012,
higashikuni_advancing_2017,abil_synthetic_2015,healy_genetic_2019,
kitada_programming_2018}.

Producing a stripe of gene expression is a fundamental patterning function
in development, and it has been shown that feed-forward motifs are minimal
networks capable of performing this function~\cite{munteanu_2014}. For example,
Cotterell and Sharpe studied what kind of 3-node network topologies could
interact with a morphogen gradient and produce a stripe of gene expression
in a one-dimensional field of cells and found a variety of networks that
implemented at least 7 different mechanisms and had different levels of
complexity, being most of them variations of feed-forward motifs~\cite{Cotterell2010}.

However, although several studies have analyzed the properties of GRNs that
process positional information, they have focused on particular natural
networks~\cite{Schaerli2014,Jaeger2004}, in synthetic networks~\cite{Basu2005,
Reinitz1995,Schaerli2018,Elowitz2000}, or have constrained the
design of the GRNs to accept an input signal from only one of its nodes, which is
not always biologically realistic as multiple input systems can exist~\cite{Cotterell2010,
munteanu_2014, Schaerli2014}.

@@

Our model is a modification of the model used by Cotterell \& Sharpe~\cite{Cotterell2010}
and proposed by Reinitz et al.~\cite{Reinitz1995}. The
model is a dynamical system that describes the change in concentration of gene
product \emph{i} in time, as shown in Eq~\ref{ec:modelo-gral}.

@@

We adopted the same definition of neighborhood as Cotterell \&
Sharpe~\cite{Cotterell2010}, in which two topologies were neighbors if the gain or
removal of any one interaction can transform one of the topologies into the
other.

@@

As a measure of network complexity, we used the complexity index based on
Shannon entropy proposed by Bonchev \& Rovray~\cite{D.2005}. We calculated this
index using the following equation:
\begin{equation}
 I_{vd} = \sum_{i=1}^V a_i \cdot \log_{2} a_i,
\end{equation}
\noindent
being $a_i$ the degree of \emph{i}-th node and $V$ the number of nodes in the
network. The network descriptors were calculated using built-in functions from
Wolfram Mathematica software.\\

@@

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network~\cite{Erdos1959}. This kind of network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}A). Nodes corresponding to topologies number one and
three showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:deg-dist}B). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
($r_S = 0.353752$, $p = 6.2933\times10^{-23} $).

@@

Besides topological robustness, we wanted to know if network topologies were
robust to perturbations in the values of the parameters. After performing the
perturbations to a GRN from each topology, we plotted the abundance of the
topology vs. its robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$)
(Fig~\ref{fig:ab-rob}A). However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif reported by Cotterell \&
Sharpe~\cite{Cotterell2010}, the I3-FFL and the ``Overlapping domains'' motif
(which is a variation of the I3-FFL) also reported by Cotterell \& Sharpe (see
Supplementary Material).

@@

\begin{table}[!ht]
% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
 \caption{{\bf Network descriptors of neutral network of topologies.}}
 \begin{tabular}{|l|l|}
 \hline
 {\bf Network descriptor} & {\bf Value}\\ \thickhline
 Average vertex degree  & 3.85714 \\ \hline
 Clustering coefficient~\cite{Watts1998} & 0.117097 \\ \hline
 Average path length    & 7.25261       \\ \hline
 Graph connectedness    & 0.00540974    \\ \hline
 Graph diameter*        & 25            \\ \hline
 Average path length*   & 9.0499        \\ \hline
 \end{tabular}
 \begin{flushleft} *These descriptors were calculated for the main connected
 subgraph of the network.
 \end{flushleft}
 \label{table1}
% \end{adjustwidth}
 \end{table}

@@

Robustness is an important feature of patterning mediated by
morphogens~\cite{lo_robust_2015}. In our study, the most abundant topologies were
found to be part of highly connected clusters in the neutral network, and this fact
indicates that these topologies are both robust to quantitative changes in the
strength of interactions and diffusion and degradation parameters as well as to
qualitative changes in the topology itself. As robustness and evolvability are
not mutually exclusive properties~\cite{wagner_robustness_2008}, it is likely
that these topologies are also evolvable.

Other studies have shown that Incoherent Feed Forward Loops are able to form
spatial stripes of gene expression~\cite{ishihara_cross_2005}. In particular,
we found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \&
Sharpe~\cite{Cotterell2010}, and has been found in transcriptional networks involved
in cell fate definition~\cite{Li2019}. Moreover, our results indicate that this
is a robust motif that can be useful for construction of synthetic gene
regulatory networks that are exposed to stochastic fluctuations in the
environment and mutations.

@@

Although it has been experimentally demonstrated that the Incoherent Feed
Forward Loop type 2 (I2-FFL) is able to generate a striped gene expression
pattern,~\cite{Schaerli2014, Basu2005}, in our study this motif was present in
only 10.92\% of the topologies. This fact may be because genetic algorithms
can not always find global minima but local minima or its surroundings, however,
it is also possible that the I2-FFL motif is not as robust as other motifs and
therefore be less represented.

We also found topologies with a similar topology to that of the “opposing
gradients” reported by Schaerli et al.~\cite{Schaerli2018,Schaerli2014}.
Although the “opposing gradients” mechanism requires the constitutive expression
of two of the genes in the GRN, and our model does not take constitutive
expression into account, we realized that the GRNs that presented the “opposing
gradients” topology, had also commonly positive auto-regulation interactions on
these two nodes, supplying in this way the lack of constitutive expression.

@@

Although GRNs that produce a band of gene expression with only two nodes have
been described (e.g. the I-zero motif,~\cite{Schaerli2014}) this is to our
knowledge the first time a GRN with a topology such as that of topology nine
(Fig \ref{fig:topol9})
and its expression mechanism are reported, increasing our knowledge of the
genotype-phenotype map in this kind of system, and it would be very interesting
to confirm experimentally the ability of this topology to generate the described
pattern and it would be interesting to study its feasibility and possible
implementation in synthetic gene regulatory networks.

@@

\paragraph*{S4 Fig.}
\label{S4_Fig}
{\bf Relevant subgraphs present in the set of GRNs.}
These are subgraphs that have been reported in previous studies as networks
involved in morphogenesis and development~\cite{Cotterell2010, Schaerli2014,
Schaerli2018}. These were used to calculate the subgraph profile and the results
reported in \nameref{S1_Table}.
